Towards a Verified Bzip2: A Functional Specification for Burrows--Wheeler

### 2019-05-05 draft_paper/draft_paper.tex
Stash

\documentclass[11pt]{thesis}

\title{Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler}
\author{Jake Waksbaum}
\date{May 6, 2019}

@@

\maketitle
\vfill

@@

The Burrows-Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. It accomplishes this by creating runs of repeated letters
in the produced string. Of course, a simple way to create runs is to
sort the string, but that would be useless for compression because the
original string would be unrecoverable.

@@

Bird and Mu \cite{birdmu,pearls} analyze the Burrows-Wheeler algorithm
in a functional setting. They express the algorithm as the composition
of many smaller functions, and then derive an expression for the
inverse transform from the specification that it inverts the forwards
transform. In addition to providing an intuitive derivation of the
inverse transform, this provides the outline of a proof of the
correctness of the Burrows-Wheeler algorithm. We give our version of
their derivation in \cref{sec:forwards_BWT,sec:inverse_BWT}.

@@

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows-Wheeler \cite{appel-func-spec}. But more
generally, this provides evidence that Bird's approach of program
calculation, in which programs are derived from their specifications
via algebraic manipulations, is useful for proving the correctness of
algorithms in a machine-checked setting. Bird has applied this
approach to greedy algorithms, the Boyer-Moore algorithm, the
Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding, and
the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
more problems \cite{pearls}.

@@

\section{Burrows-Wheeler}
\label{sec:bwt}

The Burrows-Wheeler transform was originally introduced as a pass in a
compression algorithm. First the input is transformed with BWT, which
tends to group together characters. The intuition for why this happens
is that all adjacent characters in the output string are followed in
the input string by characters that are adjacent in the sorted first
column. If there is any correlation between a character and the
character preceding it, bringing together the characters in the first
column will tend to bring together similar characters in the last
column.

@@

More precisely, suppose we had a function function \var{recreate}
such that
\begin{equation}
  \var{recreate}~(\var{bwp}~l) = \var{lexsort}~(\var{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
\begin{align}
     \label{eq:unbwt_correct_start}
     \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l} &&\\
  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l}
     &&\text{\{by \Cref{eq:recreate}\}}\\
  &= l &&\text{\{\var{recreate\_correct}\}}
     \label{eq:unbwt_correct_end}
\end{align}

How, then, do we implement \var{recreate} such that
\Cref{eq:recreate} holds? One idea is to recreate the sorted rotation
matrix column by column. We define $\var{cols}~j$ to take the first
$j$ columns of a matrix:
\lstinputlisting[
  widthgobble=2,
  firstline=19,
  lastline=20,
]{../theories/BWT/Columns.v}
Then we introduce a parameter $j$ to \var{recreate}, and our new
specification is that $\var{recreate}~j$ recreates the first $j$
columns of the sorted rotation matrix:
\begin{theorem}[recreate\_correct\_ind]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation}
    \var{recreate}~j~(\var{bwp}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
    \label{eq:recreate_correct_ind}
  \end{equation}
\end{theorem}

@@

Assuming that we can prove \var{recreate\_correct\_ind}, the
correctness of of \text{unbwt} follows:

\begin{theorem}[cols\_id]
  For all $j$ and matrices $m : \var{list}~(\var{list}~A)$,
  \begin{gather*}
    (\forall r \in m,~\var{length}~r \le n) \implies \\
    \var{cols}~n~m = m
  \end{gather*}
\end{theorem}
\begin{proof}
  $\var{cols}~j~m = \var{map}~(\var{firstn}~j)~m$, and we
  know that $m = \var{map}~(\lambda x.x)~m$. So, it suffices to show that
  for every $r \in m$, the mapping functions agree, that is
  $\var{firstn}~j~r = r$. This follows from \var{firstn\_all2}
  in the Coq standard library.
\end{proof}

\begin{theorem}[iter\_preserves]
  We are given $f$, $n$, $z$ and a predicate $P : A \to \mathbb{P}$.
  Suppose $f$ preserves $P$, which is to say that
  \begin{equation*}
    \forall x,~P~x \implies P~(f~x)
  \end{equation*}
  then,
  \begin{gather*}
     P~z \implies \forall x \in \var{iter}~f~n~z,~P~x
  \end{gather*}
\end{theorem}
\begin{proof}
  This follows from induction on $n$.
\end{proof}

\begin{lemma}[sort\_rots\_all\_len]
  For all lists $l$,
  \begin{equation*}
    \forall x \in \var{lexsort}~(\var{rots}~l),~\var{length}~x = \var{length}~l
  \end{equation*}
\end{lemma}
\begin{proof}
  Because \var{lexsort} permutes its argument, it suffices to show
  that all the rows of $\var{rots}~l$ have the same length as $l$.
  This follows from \var{iter\_preserves} and the fact that
  \var{lrot} preserves length.
\end{proof}

\begin{corollary}[recreate\_correct]
  For all lists $l$,
  \begin{equation*}
    \var{recreate}~(\var{length}~l)~(\var{bwp}~l) =
    \var{lexsort}~(\var{rots}~l)
  \end{equation*}
\end{corollary}
\begin{proof}
  This follows from \var{recreate\_correct\_ind},
  \var{cols\_id}, and \var{sort\_rots\_all\_len}.
\end{proof}

\begin{theorem}[unbwt\_correct]
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
  \end{equation*}
\end{theorem}
\begin{proof}
  The proof is almost identical to that given in
  \Crefrange{eq:unbwt_correct_start}{eq:unbwt_correct_end}.
\end{proof}

To implement \var{recreate}, we need expressions for
$\var{recreate}~0$ and $\var{recreate}~(j+1)$ that satisfy
\Cref{eq:recreate_correct_ind} in \var{recrate\_correct\_ind}:
\begin{equation}
  \var{recreate}~j~(\var{bwp}~l) =
  \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \tag{\ref{eq:recreate_correct_ind}}
\end{equation}

To derive an implementation, we can substitute for $j$ on the right-hand
side of \Cref{eq:recreate_correct_ind} and manipulate that expression
algebraically until it is in the form \var{\_ (bwp l)}. Then we can
define $\var{recreate}~0$ and $\var{recreate}~(j+1)$ to be
precisely those \var{\_}s, and by running our derivation in
reverse, we obtain a proof.

We start with the base case, $\var{recreate}~0$:
\begin{align*}
     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
  &= \var{map}~(\lambda x.[])~(\var{lexsort}~(\var{rots}~l))) \\
  &= \var{repeat}~[]~(\var{length}~(\var{lexsort}~(\var{rots}~l))) \\
  &= \var{repeat}~[]~(\var{length}~(\var{bwp}~l)) \\
     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
  &= \var{map}~(\lambda x.[])~(\var{bwp}~l)
\end{align*}

From this, we can define $\var{recreate}~0~l = \var{map}~(\lambda
x.[])~l$ and we already have the proof for the base case of the
correctness specification.

For the inductive case, we have the expression
\begin{equation*}
  \var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l))
\end{equation*}
on the right-hand side. Our plan is to express $\var{cols}~(j+1)$
in terms of $\var{cols}~j$, and then to use the induction
hypothesis
\begin{equation*}
  \var{cols}~j~(\var{lexsort}~(\var{rots}~l)) = \var{recreate}~j
\end{equation*}
to produce a well-formed recursive definition.

One way to do this is to note that taking the first $j+1$ columns of a
matrix is the same as moving the first column to the back, taking the
first $j$ columns, and then sticking the last column back on the
front:
\begin{equation*}
  \var{cols}~(j+1)~l =
  \var{prepend\_col}~(\var{map}~\var{hd}~l)~(\var{cols}~j~(\var{map}~\var{lrot}~l))
\end{equation*}
If we substitute $\var{map}~\var{rrot}~l$ for $l$ we find a simpler expression
\begin{equation}
  \label{eq:cols_rrot}
  \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) =
  \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
\end{equation}

Now we would like to find a way of introducing a $\var{map}~\var{rrot}$
before the $\var{lexsort}~(\var{rots}~l)$. The insight here is that we can
assume that \var{lexsort} is actually a radix-sort. That is, that
\begin{equation*}
\var{lexsort}~l = (\var{hdsort} \comp \var{map}~\var{rrot})^{\var{length}~l}~l
\end{equation*}
where \var{hdsort} sorts the matrix stably on the the first column.

We can do this because there is only one way to sort the matrix, so
assuming \var{lexsort} and the radix-sort are correct, they must be
doing the same thing. We will prove this in
\cref{sec:unique,sec:sort}.

Using that we can show
\begin{align*}
     \var{lexsort}~(\var{rots}~x)
  &= \var{lexsort}~(\var{rrot}~(\var{rots}~x))
     && \text{because rrot permutes its elements} \\
  &= \var{lexsort}~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
     && \text{because hdsort permutes its elements} \\
  &= (\var{hdsort} \comp \var{rrot})^n~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
     && \text{radix-sort} \\
  &= (\var{hdsort} \comp \var{rrot})^{n+1}~(\var{rots}~x)
  && \text{rep\_absorb\_r} \\
  &= \var{hdsort}~(\var{rrot}~((\var{hdsort} \comp \var{rrot})^n~(\var{rots}~x)))
     && \text{ rep\_split\_l } \\
  &= \var{hdsort}~(\var{rrot}~(\var{lexsort}~(\var{rots}~x)))
     && \text{ radix-sort }
\end{align*}

In seeking to introduce an extra \var{map rrot} term, we can note that
rotating and sorting an extra time doesn't change anything when we are
sorting the rotation matrix. This works because rotating the last
column to the front produces a matrix that is sorted on its last $n -
1$ characters, so sorting stably on the first column will re-sort the
matrix. Crucially though, rotating the last column forwards also
permutes the entire matrix, so in addition to having a sorted list, we
have a sorted list of the same elements.

More precisely, we can characterize how $\var{map}~\var{rrot}$
permutes \var{rots}.
\begin{equation*}
  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
\end{equation*}

From here we can calculate
\begin{align*}
  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
&= \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
&= \var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l))))
&&\text{ \var{hdsort} commutes with $\var{cols}~(j+1)$ } \\
&= \var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l)))
  &&\text{ reduce $\var{cols}~(j+1)$ } \\
&= \var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l)))
  &&\text{ by IH and def of bwp } \\
&= (\lambda x.~\var{hdsort}~(\var{prepend\_col}~x~(\var{recreate}~j~x)))~\var{bwp}
\end{align*}

We can use this to define \var{recreate}:
\lstinputlisting[
  widthgobble=2, firstline=156, lastline=160,
]{../theories/BWT/BurrowsWheeler.v}

@@

  If $h \not\equiv$, we again destructure $t$ and $t'$ as $l_1 \dplus h' ::
  l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
  \var{StablePerm\_destr}.
  \begin{align*}
    h::t &= h :: l_1 \dplus h' :: l_2 \\
    &\eqperm h :: h' :: l_1 \dplus l_2 \\
    &\eqperm h' :: h :: l_1 \dplus l_2 \\
    &\eqperm h' :: h :: l_1' \dplus l_2' \\
    &\eqperm h' :: l_1' \dplus h :: l_2' = h'::t
  \end{align*}
  The crucial step is $l_1 \dplus l_2 \eqperm l_1' \dplus l_2'$, which
  follows from the induction hypothesis.
\end{proof}

@@

  Using $\eqstable$ to chain \var{StablePerm}s transitively,
  \begin{align*}
    \var{apply}~(i :: p)~(a :: l)
    &= (a :: l)_i  ::
    \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
    &\eqstable (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
    &\eqstable (a :: l)_i
  \end{align*}
  where we rely on the induction hypothesis between the first and
  second lines.
\end{proof}

@@

\begin{theorem}[stable\_sort\_unique]
  \begin{equation*}
    \var{Sorted}~l \implies \var{Sorted}~l' \implies
    \var{StablePerm}~l~l' \implies l = l'.
  \end{equation*}
\end{theorem}

With \var{lexsort}, we are guaranteed stability because we are using
an order, not just a preorder, and for orders all sorts are stable.

>>>>

\documentclass[11pt, titlepage]{thesis}

\title{Towards a Verified Bzip2: A Functional Specification for Burrows--Wheeler}
\author{Jake Waksbaum}
\date{May 6, 2019}

@@

\maketitle

\tableofcontents
\clearpage

@@

The Burrows--Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. It accomplishes this by creating runs of repeated letters
in the produced string. Of course, a simple way to create runs is to
sort the string, but that would be useless for compression because the
original string would be unrecoverable.

@@

Bird and Mu \cite{birdmu,pearls} analyze the Burrows--Wheeler algorithm
in a functional setting. They express the algorithm as the composition
of many smaller functions, and then derive an expression for the
inverse transform from the specification that it inverts the forwards
transform. In addition to providing an intuitive derivation of the
inverse transform, this provides the outline of a proof of the
correctness of the Burrows--Wheeler algorithm. We give our version of
their derivation in \cref{sec:forwards_BWT,sec:inverse_BWT}.

@@

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler \cite{appel-func-spec}. But more
generally, this provides evidence that Bird's approach of program
calculation, in which programs are derived from their specifications
via algebraic manipulations, is useful for proving the correctness of
algorithms in a machine-checked setting. Bird has applied this
approach to greedy algorithms, the Boyer-Moore algorithm, the
Knuth-Morris-Pratt algoirthm, Sudoku solvers, arithmetic coding, and
the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
more problems \cite{pearls}.

@@

\section{Burrows--Wheeler}
\label{sec:bwt}

The Burrows--Wheeler transform was originally introduced as a pass in a
compression algorithm. First the input is transformed with BWT, which
tends to group together characters. The intuition for why this happens
is that all adjacent characters in the output string are followed in
the input string by characters that are adjacent in the sorted first
column. If there is any correlation between a character and the
character preceding it, bringing together the characters in the first
column will tend to bring together similar characters in the last
column.

@@

More precisely, suppose we had a function function \var{recreate}
such that
\begin{equation}
  \var{recreate}~(\var{bwp}~l) = \var{lexsort}~(\var{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
\begin{align*}
     \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l}\\
  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
  &= l
\end{align*}

How, then, do we implement \var{recreate} such that \Cref{eq:recreate}
holds? A key insight is that because any two methods of sorting the
rotation matrix must produce the same output, we can pretend that
\var{lexsort} is implemented using an sorting algorithm we find
convenient. Specifically, if we examine how an LSD radixsort sorts the
rotation matrix (\Cref{fig:radixsort_rots}) we notice an interesting
pattern. Each column of the rotation matrix is a permutation of the
original string, so when we sort a on particular column, we always
produce the same result in that column. As we move right-to-left, the
column containing the sorted input string moves with us until it
reaches the left edge of the matrix.

\begin{figure}[!hb]
  \centering
  \begin{tt}
  \setlength{\tabcolsep}{0pt}
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    abracadabra&!& \\
    bracadabra!&a& \\
    cadabra!abr&a& \\
    dabra!abrac&a& \\
    bra!abracad&a& \\
    !abracadabr&a& \\
    racadabra!a&b& \\
    ra!abracada&b& \\
    adabra!abra&c& \\
    abra!abraca&d& \\
    acadabra!ab&r& \\
    a!abracadab&r&
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    racadabra&!&ab \\
    bracadabr&a&!a \\
    acadabra!&a&br \\
    a!abracad&a&br \\
    dabra!abr&a&ca \\
    bra!abrac&a&da \\
    cadabra!a&b&ra \\
    !abracada&b&ra \\
    abra!abra&c&ad \\
    ra!abraca&d&ab \\
    abracadab&r&a! \\
    adabra!ab&r&ac
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    a&!&abracadabr \\
    r&a&!abracadab \\
    d&a&bra!abraca \\
    !&a&bracadabra \\
    r&a&cadabra!ab \\
    c&a&dabra!abra \\
    a&b&ra!abracad \\
    a&b&racadabra! \\
    a&c&adabra!abr \\
    a&d&abra!abrac \\
    b&r&a!abracada \\
    b&r&acadabra!a
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &!&abracadabra \\
    &a&!abracadabr \\
    &a&bra!abracad \\
    &a&bracadabra! \\
    &a&cadabra!abr \\
    &a&dabra!abrac \\
    &b&ra!abracada \\
    &b&racadabra!a \\
    &c&adabra!abra \\
    &d&abra!abraca \\
    &r&a!abracadab \\
    &r&acadabra!ab
  \end{tabular}
  \end{tt}
  \caption{Sorting the rotation matrix with LSD radixsort. The
    highlighted column contains the sorted original string.}
  \label{fig:radixsort_rots}
\end{figure}

We also can notice that the columns to the right of the sorted column
move with it for similar reasons. The column directly to its right is
the result of sorting the input string, and then stably sorting it
again by the character that precedes it in the original string. The
next column is the result of sorting by each character, sorting by the
preceding chracter, and then sorting by the character 2 spaces in
front.

All this means that the columns to the left of the sorted column seem
like they don't matter, which suggests that we can recreate the entire
rotation matrix from just the last column (\Cref{fig:recreate}).
However, the column direclty to the left of the sorted column is very
important, as it is the next column to be sorted. And we can see that
this column changes, only becoming the eventual last column of the
matrix in the second to last iteration. Why then, does it still work
if we use the last column everywhere?

We can think of the process of sorting the matrix as computing the
last column subject to the constraints of the lexicographic ordering.
For example, we know that the collection of all characters preceding
an \texttt{a} in the sorted column must be some permutation of
$[\texttt{c},~\texttt{d},~\texttt{r},~\texttt{r},~\texttt{!}]$,
because those are the characters that preced \texttt{a}s in the
original string. However, the order of these characters in the final
string
($[\texttt{r},~\texttt{d},~\texttt{!},~\texttt{r},~\texttt{c}]$) is
important. An \texttt{r} must be first, because in the original string
there is an \texttt{a} following an \texttt{r} which is itself
followed by a \texttt{!}, the smallest character. Similarly, although
the \texttt{d} and \texttt{!} precede \texttt{a}s that are both
followed by \texttt{bra}s, the next character for the \texttt{d} is a
\texttt{!}, and the next character for the \texttt{!} is a \texttt{c}.
This requires that the \texttt{d} preced the \texttt{!} in the last
column.

Generally, there are constraints on the relative order of the letters
in the column before the sorted column, and in each step we are
ensuring that these constraints are not violated one more character
deep in the string. By the time we have reached the last column, we
have satisfied all the constraints.

We can also think of this as computing the fixpoint of $\var{hdsort} \circ
\var{map}~\var{rrot}$: we continually move the rightmost column to the
front and sort on it, until we have sorted the whole matrix and this
becomes a no-op. The fact that the last column corresponds to the
fixpoint justifies why in \cref{fig:recreate} we can replace every
column left of the sorted column with the output of the BWT.

\begin{figure}[!hb]
  \centering
  \begin{tt}
  \setlength{\tabcolsep}{0pt}
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    \textcolor{lightgray}{aaaaaaaaaaa}&!& \\
    \textcolor{lightgray}{rrrrrrrrrrr}&a& \\
    \textcolor{lightgray}{ddddddddddd}&a& \\
    \textcolor{lightgray}{!!!!!!!!!!!}&a& \\
    \textcolor{lightgray}{rrrrrrrrrrr}&a& \\
    \textcolor{lightgray}{ccccccccccc}&a& \\
    \textcolor{lightgray}{aaaaaaaaaaa}&b& \\
    \textcolor{lightgray}{aaaaaaaaaaa}&b& \\
    \textcolor{lightgray}{aaaaaaaaaaa}&c& \\
    \textcolor{lightgray}{aaaaaaaaaaa}&d& \\
    \textcolor{lightgray}{bbbbbbbbbbb}&r& \\
    \textcolor{lightgray}{bbbbbbbbbbb}&r&
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    \textcolor{lightgray}{aaaaaaaaa}&!&ab \\
    \textcolor{lightgray}{rrrrrrrrr}&a&!a \\
    \textcolor{lightgray}{ddddddddd}&a&br \\
    \textcolor{lightgray}{!!!!!!!!!}&a&br \\
    \textcolor{lightgray}{rrrrrrrrr}&a&ca \\
    \textcolor{lightgray}{ccccccccc}&a&da \\
    \textcolor{lightgray}{aaaaaaaaa}&b&ra \\
    \textcolor{lightgray}{aaaaaaaaa}&b&ra \\
    \textcolor{lightgray}{aaaaaaaaa}&c&ad \\
    \textcolor{lightgray}{aaaaaaaaa}&d&ab \\
    \textcolor{lightgray}{bbbbbbbbb}&r&a! \\
    \textcolor{lightgray}{bbbbbbbbb}&r&ac
  \end{tabular}
  $\rightarrow \cdots \rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    \textcolor{lightgray}{a}&!&abracadabr \\
    \textcolor{lightgray}{r}&a&!abracadab \\
    \textcolor{lightgray}{d}&a&bra!abraca \\
    \textcolor{lightgray}{!}&a&bracadabra \\
    \textcolor{lightgray}{r}&a&cadabra!ab \\
    \textcolor{lightgray}{c}&a&dabra!abra \\
    \textcolor{lightgray}{a}&b&ra!abracad \\
    \textcolor{lightgray}{a}&b&racadabra! \\
    \textcolor{lightgray}{a}&c&adabra!abr \\
    \textcolor{lightgray}{a}&d&abra!abrac \\
    \textcolor{lightgray}{b}&r&a!abracada \\
    \textcolor{lightgray}{b}&r&acadabra!a
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &!&abracadabra \\
    &a&!abracadabr \\
    &a&bra!abracad \\
    &a&bracadabra! \\
    &a&cadabra!abr \\
    &a&dabra!abrac \\
    &b&ra!abracada \\
    &b&racadabra!a \\
    &c&adabra!abra \\
    &d&abra!abraca \\
    &r&a!abracadab \\
    &r&acadabra!ab
  \end{tabular}
  \end{tt}
  \caption{Recreating the rotation matrix from the last column.}
  \label{fig:recreate}
\end{figure}

We implement \var{recreate} as follows, introducing a parameter $j$ so
that we can recursively recreate the first $j$ columns of the matrix,
one column at a time. Here $l$ is meant to be the output of the BWT,
the last column of the sorted rotation matrix.
\lstinputlisting[
  widthgobble=2, firstline=156, lastline=160,
]{../theories/BWT/BurrowsWheeler.v}
This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
\lstinputlisting[
  widthgobble=2, firstline=47, lastline=48,
]{../theories/BWT/Columns.v}
\lstinputlisting[
  widthgobble=2, firstline=16, lastline=20,
]{../theories/BWT/Lib/ZipWith.v}

@@

Since we recreate the matrix column by column, we can define an
invariant using $\var{cols}~j$, which takes the first $j$ columns of a matrix.
\lstinputlisting[
  widthgobble=2,
  firstline=19,
  lastline=20,
]{../theories/BWT/Columns.v}

\begin{theorem}[recreate\_correct\_inv]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation*}
    \var{recreate}~j~(\var{bwp}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \end{equation*}
\end{theorem}

Assuming we can prove this theorem, we will be able to prove that our
implementation of \var{unbwt} is correct along the same lines as before.

To prove this invariant, we need to prove a few other theorems that
have already come up when explaining the algorithm conceptually.
First, we note that moving the last column of the rotation matrix to the front
is the same as moving the last row to the top, because the rotation
matrix is symmetric.

\begin{theorem}[map\_rrot\_rots]
  For all lists $l$,
  \begin{equation*}
    \var{map}~\var{rrot}~(\var{rots}~l) = \var{rrot}~(\var{rots}~l)
  \end{equation*}
\end{theorem}
\begin{proof}
  Letting $n = \var{length}~l$
  \begin{align*}
       &\var{map}~\var{rrot}~(\var{rots}~l) \\
    ={}&\var{map}~\var{rrot}~[l,~\var{lrot}~l,~\var{lrot}^2~l,~\ldots,~\var{lrot}^{n-1}~l] \\
    ={}&[\var{rrot}~l,~l,~\var{lrot}~l,~\ldots,~\var{lrot}^{n-2}~l] \\
    ={}&[\var{lrot}^{n-1}~l,~l,~\var{lrot}~l,~\ldots,~\var{lrot}^{n-2}~l] \\
    ={}&\var{rrot}~[l,~\var{lrot}~l,~\ldots,~\var{lrot}^{n-1}~l] \\
    ={}&\var{rrot}~(\var{rots}~l) \qedhere
  \end{align*}
\end{proof}

Now we can prove that the sorted rotation matrix is indeed a fixpoint.
\begin{theorem}[lexsort\_rots\_hdsort]
  For all lists $l$,
  \begin{equation*}
    (\var{hdsort} \circ
    \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) =
    \var{lexsort}~(\var{rots}~l)
  \end{equation*}
\end{theorem}
\begin{proof}
  By \var{StablePerm\_Sorted\_eq}, proved in \cref{sec:unique}, any
  two sorts are equivalent, and we can replace \var{lexsort} with an
  LSD radixsort implemented as
  \begin{equation*}
    \var{radixsort}~m \coloneqq (\var{hdsort} \circ \var{map}~\var{rrot})^n~m
  \end{equation*}
  where $m$ is a matrix with $n$ columns. In this case, we are sorting
  the rotation matrix and we let $n = \var{length}~l$. Then we have
  \begin{align*}
       &(\var{hdsort} \circ \var{map}~\var{rrot})~(\var{lexsort}~(\var{rots}~l)) \\
    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})~((\var{hdsort} \circ \var{map}~\var{rrot})^n~(\var{rots}~l)) \\
    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})^{n+1}~(\var{rots}~l) \\
    ={}&(\var{hdsort} \circ \var{map}~\var{rrot})^n~((\var{hdsort} \circ \var{map}~\var{rrot})~(\var{rots}~l)) \\
    ={}&\var{lexsort}~((\var{hdsort} \circ \var{map}~\var{rrot})~(\var{rots}~l)) \\
    ={}&\var{lexsort}~((\var{hdsort} \circ \var{rrot})~(\var{rots}~l)) \\
    ={}&\var{lexsort}~(\var{rots}~l)
  \end{align*}
  The last step follows because the function $\var{hdsort} \circ
  \var{rrot}$ permutes the rotation matrix
\end{proof}

We need two more lemmas that describe the interaction between
\var{cols} and $\var{map}~\var{rrot}$.

\begin{theorem}[cols\_hdsort\_comm]
  For all matrices $m$ and naturals $j$,
  \begin{equation*}
    \var{cols}~(j+1)~(\var{hdsort}~m) = \var{hdsort}~(\var{cols}~(j+1)~m)
  \end{equation*}
\end{theorem}
\begin{proof}
  \var{hdsort} only examines the first elements of the rows of $m$,
  and $\var{cols}~(j+1)$ doesn't alter the first elements of the rows
  of $m$.\footnote{This argument is made precise by
    \var{key\_sort\_inv} in \cref{sec:lex_ord}.}
\end{proof}

\begin{theorem}[cols\_map\_rrot]
  For all $j$ and matrices $m$ that has rows all of length less than $j$,
  \begin{equation*}
    \var{cols}~(j+1)~(\var{map}~\var{rrot}~m) =
    \var{prepend\_col}~(\var{map}~\var{last}~m)~(\var{cols}~j~m)
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by induction on $m$. In the inductive case we need to show that
  \begin{equation*}
    \var{cols}~(j+1)~(\var{map}~\var{rrot}~(r::m)) =
    \var{prepend\_col}~(\var{map}~\var{last}~(r::m))~(\var{cols}~j~(r::m))
  \end{equation*}
  Simplifying, we have that the tails are equal by the induction
  hypothesis, and we must show that the heads are equal.
  \begin{align*}
    \var{firstn}~(j+1)~(\var{rrot}~r) &= \var{last}~r :: \var{firstn}~j~r \\
    \var{firstn}~(j+1)~(\var{last}~r :: \var{init}~r) &= \var{last}~r
    :: \var{firstn}~j~r \\
    \var{last}~r :: \var{firstn}~j~(\var{init}~r) &= \var{last}~r :: \var{firstn}~j~r \\
    \var{firstn}~j~(\var{init}~r) &= \var{firstn}~j~r
  \end{align*}
  The bottom equation is true by another theorem we will not prove here.
\end{proof}

Now we are ready to prove the invariant:
\begin{theorem}[recreate\_correct\_inv]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation*}
    \var{recreate}~j~(\var{bwp}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \end{equation*}
\end{theorem}
\begin{proof}
\begin{align*}
  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
={}& \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
={}&
\var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
={}&
\var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l))) \\
={}&
\var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l))) \\
={}& \var{recreate}~j~(\var{bwp}~l) \qedhere
\end{align*}
\end{proof}

@@

  If $h \not\equiv$, we again destructure $t$ and $t'$ as $l_1 \dplus h' ::
  l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
  \var{StablePerm\_destr}.
  \begin{align*}
    h::t ={}& h :: l_1 \dplus h' :: l_2 \\
    \eqperm{}& h :: h' :: l_1 \dplus l_2 \\
    \eqperm{}& h' :: h :: l_1 \dplus l_2 \\
    \eqperm{}& h' :: h :: l_1' \dplus l_2' \\
    \eqperm{}& h' :: l_1' \dplus h :: l_2' \\
    ={}& h'::t
  \end{align*}
  The crucial step is $l_1 \dplus l_2 \eqperm l_1' \dplus l_2'$, which
  follows from the induction hypothesis.
\end{proof}

@@

  Using $\eqstable$ to chain \var{StablePerm}s transitively,
  \begin{align*}
    \var{apply}~(i :: p)~(a :: l)
    ={}& (a :: l)_i  ::
    \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
    \eqstable{}& (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
    \eqstable{}& (a :: l)_i
  \end{align*}
  where we rely on the induction hypothesis between the first and
  second lines.
\end{proof}

@@

\begin{theorem}[StablePerm\_Sorted\_eq]
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \implies \var{Sorted}~l' \implies
    \var{StablePerm}~l~l' \implies l = l'.
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by induction on $l$. In the inductive case, we have to
  show that
  \begin{equation*}
    h :: t = h' :: t'
  \end{equation*}
  given $\var{Sorted}~(h :: t)$, $\var{Sorted}~(h' :: t')$, and
  $\var{StablePerm}~(h :: t)~(h' :: t')$.
